---
title: "fall-bloomers"
author: "jt-miller"
date: "2024-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fall Bloomers 
The initial sliding window analysis indicated some interesting trends on the shifts in the fall. At the community level, there seems to be a shift towards later flowering periods. This is a relatively unexplored topic in desert phenology and could have significant implications on these taxa's ecology. <br>

Here, I will pull out the taxa that can be sufficiently described as fall blooming taxa by the data and assess species level variation in these phenological trends. 

Load Libraries
```{r}
library(data.table)
library(tidyverse)
library(kde1d)
library(stringr)
```

### Fall subsetting 
Fall is arbitrary in terms of when it actually begins, but there is a distinct period for when early and later (in doy) flowering events occur. Looking at my previous analysis, for plants it appears there is a local minimum ~200 doy. As a first run we will define fall flowering plants as those that bloom >= 200 doy. <br>
After a closer look at the taxonomy used, its apparent that there are sub-species level taxonomic designations within the dataset. Unfortunately there really isn't a way to determine the validity of these without expert opinion, so Im going to collapse these to the species level as a conservative approach. 
```{r}
hpg <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-plant-data.csv") # load in high coverage flowering plants


hpg <- hpg %>% 
  mutate(speciesName = word(scientificName, 1,2))
```
The first question then is what taxa (that have sufficient data) are considered spring or fall (or both) blooming plants?
```{r}

hpg$season <- ifelse(hpg$doy >= 200, "fall", "spring")
plant_seasons <- hpg[, seasonsPresent := { # creates a new field called seasonPresent
    if ("spring" %in% season & "fall" %in% season) { # if both spring and fall or present for a species 
      as.factor("both") # call it both
    } else if ("spring" %in% season) { # else if spring is present
      as.factor("spring") # call it spring 
    } else if ("fall" %in% season) { # else if fall is present
      as.factor("fall") # call it fall
    } else NA_integer_ # else it is NA
}, by = speciesName] # group by speciesName 


# Build the ten year bins 
grid_holder_p <- list()
gid_p_v <- unique(plant_seasons$GID)
for(i in 1:length(unique(plant_seasons$GID))){
  grid_holder_p[[i]] <- plant_seasons[GID == gid_p_v[[i]],]
}

# Function that uses a for-loop to create the decadal subsets
decadal_subsetter <- function(dataset, start_year, end_year, bin_size){
  decadal_subsets <- list() # empty storage list
for(i in start_year:(end_year - bin_size)){
  end <- i + bin_size # Our end condition 
  subset_data <- dataset[year >= i & year <= end, ] # Subsets the data into bins on a decadal sliding scale
  decadal_subsets[[paste(i, end, sep = "-")]] <- subset_data # Label
}
  return(decadal_subsets)
}

# Set arguments
start_year <- 1950
end_year <- 2023
bin_size <- 10
gids <- c(217, 229, 230, 241)


nested_list_holder_p <- list()
for(i in 1:length(grid_holder_p)){
decadal_subsets <- decadal_subsetter(dataset = grid_holder_p[[i]], # Take the ith grid and do the decadal subsetting  
                  start_year = start_year, 
                  end_year = end_year, 
                  bin_size = bin_size)
nested_list_holder_p[[paste(gids[i], "grid", sep = "-")]] <- decadal_subsets
}

# Bring lists together
table_1 <- rbindlist(nested_list_holder_p[[1]], idcol = TRUE) # one dataframe to rule them all
table_2 <- rbindlist(nested_list_holder_p[[2]], idcol = TRUE)
table_3 <- rbindlist(nested_list_holder_p[[3]], idcol = TRUE)
table_4 <- rbindlist(nested_list_holder_p[[4]], idcol = TRUE)
full_table <- rbind(table_1, table_2, table_3, table_4) # and one to bind them

# Find the representation of seasons among the time bins 
table_seasons <- full_table[, uniqFallDOYobs := sum(!duplicated(doy[doy >= 200])), by = .(speciesName, .id)] # sum up unique doy equal to or past 200
table_seasons <- table_seasons[, uniqSpringDOYobs := sum(!duplicated(doy[doy < 200])), by = .(speciesName, .id)] # sum up unique doy less than 200

plant_blooming_period <- table_seasons[, bloom_period := ifelse(uniqSpringDOYobs >= 5 & uniqFallDOYobs >= 5, "opportunistic bloomer",
                                       ifelse(uniqSpringDOYobs >= 5 & uniqFallDOYobs < 5, "spring bloomer",
                                              ifelse(uniqSpringDOYobs < 5 & uniqFallDOYobs >= 5, "fall bloomer",
                                                     "unknown"))), by = .(speciesName, .id)]
```
Fall & Opportunistic Bloomers <br> 
Opportunistic bloomers present two interesting scenarious: Either they are as described and have liable flowering periods that will just bloom during any 'optimal' conditions regardless of temporal constraints OR they could also possibly be an artifact of our abstraction 'fall' and 'spring' e.g. if a species of plant that typically flowers late fall (winter) is pushed into Jan then it would be considered a spring bloomer (still winter really) <br> 
```{r}
fallish_plants <- plant_blooming_period[, speciesBloomsFall := ifelse(any(bloom_period %in% c("opportunistic bloomer", "fall bloomer")), TRUE, FALSE), by = .(speciesName)] # just by species since we want to know if it was ever a fall/opportunistic bloomer. 

fallish_plants <- fallish_plants[speciesBloomsFall == TRUE] # grab only those records that are fall bloomers or are partially. 

fallish_plants <- setnames(fallish_plants, old = ".id", new = "time_bin")

fwrite(fallish_plants, "/home/jt-miller/Gurlab/sliding-phenology/data/processed/fallish-plants.csv")
```


### kernel density estimations 
Create kernel density estimations to see if we can recover those shifts in fall blooms per species
```{r}
dt_long <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/fallish-plants.csv")

# Reorganize into nested lists for the loops  
# Geographic Binning 
gid_p_v <- unique(dt_long$GID) # create a vector for later
grid_holder_p <- split(dt_long, dt_long$GID) # split into unique GID lists

# Temporal Subsetting 
time_bins <- list()
for(i in 1:length(grid_holder_p)){
time_bins[[i]] <- split(grid_holder_p[[i]], grid_holder_p[[i]]$time_bin)
}

# Sp level kde simus 
nested_list_holder_p <- time_bins # rename the var 

for(w in 1:length(nested_list_holder_p)){
  spatial_ID <- names(nested_list_holder_p[w]) # replace with w
  spatial_bin <- nested_list_holder_p[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$speciesName) # extract a vector of uniq scientific names from bin
    for(i in 1:length(unique_names_in_bin_v)){
      named_subset <- temporal_spatial_bin[speciesName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(named_subset) <= 5 | length(unique(named_subset$doy)) <= 3){ # adjust to 3 days***
    named_phenometric_summary <- data.table(taxa = "plant", speciesName = unique(named_subset$speciesName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    named_subset <- named_subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = named_subset$doy) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    name <- gsub(",", "", unique(named_subset$speciesName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/fall-plots/", "plant_", name, "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0(unique(named_subset$speciesName), " kernel density distribution for time bin ", spatial_temporal_ID), xlab = "day of year", xlim = c(0,365))
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(taxa = "plant", speciesName = unique(named_subset$speciesName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/fall-tables/", "plant_", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(named_subset$speciesName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/fall-tables/", "plant_", name, "_", spatial_temporal_ID, ".csv"))
    } # end of i loop
  } # end of j loop
} # end of w loop
```
### Reproduce the community level trends with the fall(ish) plants. 
```{r}
# Run the nested loop that runs kernel denisty on community level phenology
for(w in 1:length(nested_list_holder_p)){
  spatial_ID <- names(nested_list_holder_p[w]) # replace with w
  spatial_bin <- nested_list_holder_p[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$speciesName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
    #for(i in 1:length(unique_names_in_bin_v)){
      #named_subset <- temporal_spatial_bin[speciesName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(subset) <= 5 | length(unique(subset$doy)) <= 3){
    named_phenometric_summary <- data.table(taxa = "plants", GID = unique(subset$GID), number_of_records = nrow(subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density. xmin = 0, xmax = 365
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde-fall-community-plots/", "plants", "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0("Plant Community ", " Kernel Density Distribution for Time Bin ", spatial_temporal_ID), xlab = "day of year", xlim = c(0,365))
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(taxa = "plants", GID = unique(subset$GID),
                                            number_of_records = nrow(subset), time_bin = named_temporal_bin)
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
    named_phenometric_summary <- named_phenometric_summary[, num_of_species := num_of_species]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde-fall-community-tables/", "plants", "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(subset$speciesName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde-fall-community-tables/", "plants", "_", spatial_temporal_ID, ".csv"))
    #} # end of i loop
  } # end of j loop
} # end of w loop


# Community level trends on the same plots 
# Create an empty storage df  
result_df <- data.frame()
nested_kde_holder <- list()

for(w in 1:length(nested_list_holder_p)){
  spatial_bin <- nested_list_holder_p[[w]] # replace with w 
  spatial_ID <- names(nested_list_holder_p[[1]][1])
  kde_holder <- list() # intialize a ist to store kde sims for each w
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$speciesName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
       if(nrow(subset) <= 5 | length(unique(subset$doy)) <= 3){
         print("Skipping Plot Due to Insufficient Data")
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density. xmin = 0, xmax = 365
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Store data 
    row_data <- data.frame(spatialTemporalID = spatial_temporal_ID, numberSpecies = num_of_species, quant50=quant50, var=var50)
    result_df <- rbind(result_df, row_data)
    # Store kde.sims in nested structure
      kde_holder[[named_temporal_bin]] <- list(
        quant50 = quant50,
        var50 = var50,
        kde.simulations = kde.sims)
    

} # End of else statement
    
  } # end of j loop
  
  nested_kde_holder[[spatial_ID]] <- kde_holder # Store the kde_holder for the current 'w' in nested_kde_holder
} # end of w loop

# Modified code to include legends with years ########################################################################################################
library(RColorBrewer)
png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/graphical-outputs/", "grid-cell-241-sample-dists-fall", ".png"), width = 800, height = 600)
plot(1, type = "n", main = "Plant Community Kernel Density Distribution by Decadal Bin for Grid Cell 241 Fall", xlab = "Decadal Bin", ylab = "Density", xlim = c(0, 400), ylim = c(0, 0.015))

# Extract the first 5 and last 5 time bins
cell_241 <- nested_kde_holder[[3]]
time_bins_to_plot <- c(cell_241[1:5], cell_241[(length(cell_241)-4):length(cell_241)])

# Use a color palette for lines
#line_colors <- c("lightblue", "lightgreen", "lightorange", "lightpurple", "lightred", "darkblue", "darkgreen", "darkorange", "darkpurple", "darkred")

light_colors <- brewer.pal(5, "Set3")
dark_colors <- brewer.pal(5, "Dark2")
line_colors <- c(light_colors, dark_colors)
# Create a vector to store legend labels
years <- names(time_bins_to_plot)
legend_labels <- c()
# Now plot these kernel densities iteratively 
for (w in 1:length(time_bins_to_plot)) {
  spatial_id <- names(time_bins_to_plot[w]) # replace with w
  spatial_bin <- time_bins_to_plot[[w]] # replace with w
  year <-  years[w] # Assuming the year information is stored in the first element

  legend_labels[w] <- paste("Kernel Density (", year, ")", sep = "")

  for (j in 1:length(spatial_bin)) {
    quant50 <- spatial_bin[1][[1]]
    kde_sim <- spatial_bin[[3]]
    var50 <- spatial_bin[[2]]
    
    # Plot kernel density using lines with different colors
    lines(kde_sim, col = line_colors[w], xaxt = "n")

    # Add vertical line for 50th percentile
    abline(v = quant50, col = line_colors[w], lty = 2)

    # Add error bars
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    segments(x0 = quant50, y0 = lower_bound, x1 = quant50, y1 = upper_bound, col = line_colors[w], lwd = 2)
  }
} # end of w loop.

# Add legend outside the loop to avoid duplication
legend("topright", legend = legend_labels[1:10], col = line_colors[1:length(time_bins_to_plot)], lty = 1, lwd = 1, cex = 0.75)
dev.off()
```



From just some playing around with the numbers, an interesting fall bloomer is Eriogonum wrightii, wrights buckwheat. It appears to be a fall bloomer that is occasionally pushed into the next year. Taxonomically I think I could also collapse it so that all the varities are included in as one species, giving appropriate coverage. 



### As to the question about whether fall plant phenology is changing...I see three explanations/hypotheses for why this could occur.
H1: Plants that flower in the fall are shifting to later doy flowering due to some change in abiotic cues. 
H2: Plants that are capable of flowering multiple times per year are starting to flower in the fall, shifting the average. 
H3: The 'shift' is an artifact of the data. The early periods had such low taxonomic representation that comparing the modern and 1950s community phenology does not reveal any true biological signal. 

First, it would be informative to rule out H3 as that explanation suggests that the trends we are seeing are an artifact of how the data was collected rather than the actual underlying biology. 
### Find some well represented indicator species to dive deeper into
```{r}
# Initialize empty list to store results
results <- list()

input_dir <- "/home/jt-miller/Gurlab/sliding-phenology/outputs/fall-tables//"
# Initialize empty list to store results
results <- list()

# Loop over all png files in the input directory
output_table <- data.table()
files <- dir(input_dir, full.names = TRUE, pattern = ".csv")
for (i in 1:length(files)) {
  table <- fread(files[[i]])
  pattern <-  "(\\d{4}-\\d{4})" # Create a pattern for the date
  date <- str_extract(files[[i]], pattern)
  if(is.na(table$fiftieth_percentile) == TRUE | is.na(table$speciesName == TRUE)){
    next
  } else{
     output_table <- rbindlist(list(output_table, data.table(speciesName = table$speciesName, fiftieth_percentile = table$fiftieth_percentile, date = date, GID = table$GID)))
  }
}

#output_table <- do.call("rbind", output_table)

output_df <- as.data.frame(output_table)
# Count total number of observations per species
sampling <- output_df %>%
  filter(!is.na(speciesName)) %>% 
  group_by(speciesName, GID) %>%
  summarize(unique_time_bins = length(unique(date))) %>% 
  arrange(desc(unique_time_bins))

# Print resulting tally table
print(output_table)

```