---
title: "pulling-phylogeny"
author: "jt-miller"
date: "2024-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Some code to look at the overall phylogenetic placement of the taxa we are modeling. 
Previous literature has identified that overall trends in phenology are thought to be shared across the phylogeny for plants (i.e. evolutionary history will influence flowering). We can both:
1) Test this hypothesis in my datasets.
2) Evaluate the overall scope of the tree that we are capturing in this analysis. 

Load in Packages 
```{r}
library(rotl) # calls the open tree api to give us a rough idea of the current phylogeny, I would still take these with a grain of salt but its not a bad exploratory option
library(data.table)
library(ggplot2)
library(stringr)
library(tidyverse)
library(phytools)
library(ape)
```
Load Data
```{r eval=FALSE, include=FALSE}
#data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-plant-data.csv")
#fall_bloomers <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/fallish-plants.csv")
data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/blooming-period-dataset.csv")
```

Simplify to just names and create a simple subset for illustration purposes 
```{r eval=FALSE, include=FALSE}
#data <- mutate(data, genusName = word(speciesName, 1)) # grab out the resolved genus
names <- unique(data$scientificName)
names <- unique(data$speciesName)
name_subset <- names
resolved_names <- tnrs_match_names(name_subset) # uses tnrs tool to match up names to a ott_id
resolved_names2 <- filter(resolved_names, !is.na(unique_name))
resolved_lineages <- tax_lineage(taxonomy_taxon_info(ott_id(resolved_names), include_lineage = TRUE))
resolved_lineages2 <- tax_lineage(taxonomy_taxon_info(ott_id(resolved_names2), include_lineage = TRUE))
df <- data.frame()
for(i in 1:length(resolved_lineages)){
  temp_df <- data.frame(rank = resolved_lineages[[i]][1], name = resolved_lineages[[i]][2], ott_id = resolved_lineages[[i]][4])
  temp_df <- as.data.frame(temp_df)
  df <- rbind(df, temp_df)
}

genus_df <- subset(df, df$rank == "genus")
genus_df <- distinct(genus_df, rank, name, ott_id)
```

Grab a tree for these tips 
```{r eval=FALSE, fig.height=95, fig.width=25, include=FALSE}

#tr <- tol_induced_subtree(ott_ids = ott_id(resolved_names))
tr <- tol_induced_subtree(ott_ids = genus_df$ott_id)
tr <- tol_induced_subtree(ott_ids = ott_id(resolved_names2))
# Set the size of the plotting device
#phylo_tree <- as.phylo(tr)
#tree_structure <- tr$edge
#phylo_tree <- ape::phylo(tree_structure)
#par(pin = c(5, 5))  # Adjust the values according to your preferences
#par(mar = c(5, 4, 4, 2) + 0.1)
#plot(tr, )
#ggtree(tr, layout="rectangular", open.angle=10, ggs = list(width = 20, height = 15)) +
  #geom_tiplab()
par(mar = c(10, 10, 10, 2)) 
ape::plot.phylo(tr, cex = 1.5, type = "phylogram")
  
```
### Labeling Phylogeny
In order to address both (1) and (2), it may be helpful to label the phylogeny as a overall visual. 
```{r eval=FALSE, include=FALSE}
named_data_labeled <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/blooming-period-dataset.csv") # see fall-bloomers.Rmd for code that denotes what seasonal bloom period a species is assigned. 
names <- unique(named_data_labeled$speciesName) # just need the names
name_subset <- names# for testing purposes only, ignore as a variable change if no subset applied
resolved_names <- tnrs_match_names(name_subset) # uses tnrs tool to match up names to a ott_id 
named_data_labeled_d <- distinct(named_data_labeled, speciesName, scientificName, bloom_period) # take only distinct values

scope_labels <- named_data_labeled_d  %>%
  filter(!bloom_period == "unknown") %>%  # doesnt contribute to known phenology
  group_by(speciesName) %>%
  summarise(bloomSpecificity = case_when(
    n_distinct(bloom_period) == 1 & bloom_period == 'spring bloomer' ~ "spring bloomer",
    n_distinct(bloom_period) == 1 & bloom_period == 'fall bloomer' ~ "fall bloomer",
     n_distinct(bloom_period) > 1 ~ "opportunistic bloomer"
  )) %>% 
  distinct(speciesName, bloomSpecificity)

scope_labels <- named_data_labeled_d  %>%
  filter(!bloom_period == "unknown") %>%  # doesnt contribute to known phenology
  group_by(scientificName) %>%
  mutate(scientificNameBloomSpecificity = case_when(
    n_distinct(bloom_period) == 1 & bloom_period == 'spring bloomer' ~ "spring bloomer",
    n_distinct(bloom_period) == 1 & bloom_period == 'fall bloomer' ~ "fall bloomer",
     n_distinct(bloom_period) > 1 ~ "opportunistic bloomer"
  )) %>% 
  ungroup() %>% 
  group_by(speciesName) %>% 
  mutate(speciesNameBloomSpecificity = case_when(
    n_distinct(bloom_period) == 1 & bloom_period == 'spring bloomer' ~ "spring bloomer",
    n_distinct(bloom_period) == 1 & bloom_period == 'fall bloomer' ~ "fall bloomer",
     n_distinct(bloom_period) > 1 ~ "opportunistic bloomer"
  )) %>% 
  distinct(scientificName, speciesName, scientificNameBloomSpecificity, speciesNameBloomSpecificity)

# for now, lets just go with the scientificName level of the field 
scope_labels <- scope_labels %>% 
  distinct(scientificName, scientificNameBloomSpecificity)

specific_bloomers <- scope_labels %>% 
  filter(scientificNameBloomSpecificity %in% c("fall bloomer", "spring bloomer"))

name_subset <- specific_bloomers$scientificName

resolved_subset <- tnrs_match_names(name_subset)

in_tree <- is_in_tree(ott_id(resolved_subset))
in_tree
# resolved_subset <- filter(resolved_subset, !is.na(unique_name))
# resolved_subset <- filter(resolved_subset, !flags == "hidden")
# resolved_subset <- filter(resolved_subset, !flags == "sibling_higher")
resolved_subset2 <- resolved_subset %>% 
  filter(!grepl("incertae_sedis", flags)) %>% 
  #filter(!unique_name == "Parkinsonia (genus in kingdom Archaeplastida)") %>% 
  filter(!grepl("Parkinsonia", unique_name)) %>% 
  #filter(!unique_name == "Tidestromia oblongifolia") %>%
  filter(!grepl("Tidestromia", unique_name)) %>% 
  #filter(!unique_name == "Salvia vaseyi") %>%
  filter(!grepl("Salvia", unique_name)) %>% 
  #filter(!unique_name == "Salvia carduacea") %>%
  #filter(!unique_name == "Camissonia brevipes") %>%
  filter(!grepl("Camissonia", unique_name)) %>%
  #filter(!unique_name == "Ceanothus perplexans") %>%
  filter(!grepl("Ceanothus", unique_name)) %>%
  #filter(!unique_name == "Sphaeralcea") %>%
  filter(!grepl("Sphaeralcea", unique_name)) %>%
  #filter(!unique_name == "Sorghum") %>%
  filter(!grepl("Sorghum", unique_name)) %>%
  #filter(!unique_name == "Suaeda") %>%
  filter(!grepl("Suaeda", unique_name)) %>%
 #filter(!unique_name == "Camissonia pallida") %>%
  #filter(!unique_name == "Stipa") %>%
  filter(!grepl("Stipa", unique_name)) %>%
 # filter(!unique_name == "Chenopodium") %>%
  filter(!grepl("Chenopodium", unique_name)) %>%
  #filter(!unique_name == "Camissonia cardiophylla") %>%
  #filter(!unique_name == "Chaenactis") %>%
  filter(!grepl("Chaenactis", unique_name)) %>%
 # filter(!unique_name == "Sarcostemma") %>%
  filter(!grepl("Sarcostemma", unique_name)) %>%
  #filter(!unique_name == "Galium") %>%
  filter(!grepl("Galium", unique_name)) %>%
 # filter(!unique_name == "Baileya (genus in subkingdom SAR)") %>%
  filter(!grepl("Baileya", unique_name)) %>%
  #filter(!unique_name == "Coreopsis") %>%
  filter(!grepl("Coreopsis", unique_name)) %>%
  #filter(!unique_name == "Muhlenbergia")
filter(!grepl("Muhlenbergia", unique_name))

in_tree <- is_in_tree(ott_id(resolved_subset2))
in_tree

tree_object <- tol_induced_subtree(ott_ids = ott_id(resolved_subset))
  

# remove corrupting branches...?
resolved_subset <- resolved_subset %>% 
  filter(!grepl("Baileya", unique_name)) %>% 
  filter(!grepl("Funastrum", unique_name)) %>% 
  filter(!grepl("Funastrum", unique_name)) %>% 
  filter(!grepl("Salvia", unique_name)) %>% 
  filter(!grepl("Camissonia", unique_name))
# resolved_subset <- filter(resolved_subset, number_matches == 1)
# in_tree <- list()
# for(i in 1:length(resolved_subset$search_string)){
#   in_tree[[i]] <- is_in_tree(ott_id(subset(resolved_subset, resolved_subset$search_string == resolved_subset$search_string[[i]])))
# }
tree_object <- tol_induced_subtree(ott_ids = ott_id(resolved_subset))
# in_tree <- resolved_subset %>% 
#   mutate(in_tree = is_in_tree(ott_id(resolved_subset)))
resolved_names_m <- merge(resolved_names, scope_labels, by.x = "unique_name", by.y = "speciesName")
resolved_names_m <- merge(resolved_subset, scope_labels, by.x = "unique_name", by.y = "speciesName")
tree_object <- tol_induced_subtree(ott_ids = ott_id(resolved_names)) # appends lineages from open tree

#merge(tree_object, names, by = "scientificName")
# arranged <- resolved_names_m %>% 
#   arrange(search_string)
#bloom <- as.factor(resolved_names_m$bloomSpecificity)
#names(bloom) <- tree_object$tip.label 

#tree_obj_merge <- merge(tree_object)

```

Create some labeled plots 
```{r eval=FALSE, include=FALSE}
ott_ids <- tree_object[[2]]
ott_ids <- gsub("[^0-9 ]", "", ott_ids)
ott_names <- gsub("\\d+", "", tree_object[[2]])
ott_names <- gsub("\\_ott", "", ott_names)
ott_names <- gsub("_", " ", ott_names)
tree_df <- data.frame(ott_ids, ott_names)
trait_df <- merge(tree_df, resolved_names_m, by.x = "ott_names", by.y = "unique_name", all = TRUE)
trait_df <- select(trait_df, ott_id, ott_names, bloomSpecificity)
trait_df <- trait_df %>% 
  mutate(bloomCol = case_when(
    is.na(bloomSpecificity) ~ "grey", 
    bloomSpecificity == "spring bloomer" ~ "darkgreen", 
    bloomSpecificity == "fall bloomer" ~ "darkorange", 
    bloomSpecificity == "opportunistic bloomer" ~ "goldenrod"
    
  ))

# X <- c("green", "orange", "yellow") 
# ape::plot.phylo(tree_object, cex = 1.5, type = "phylogram", edge.color = sample(X, length(tree_object$edge), replace = TRUE),
#      edge.width = rep(5, length(tree_object$edge)))

ape::plot.phylo(tree_object, cex = 1.5, type = "phylogram", tip.color = trait_df$bloomCol) # edges are being a pain...
```
New plan, new plan: Find representative taxa 
1) That have the best amount of temporal coverage for year by year phenometric measurements
```{r}
named_data_labeled <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/blooming-period-dataset.csv") 
# 
flowering_data <- named_data_labeled


result_table <- flowering_data %>%
  group_by(speciesName, year) %>%
  summarise(
    condition_satisfied = n() > 5 & n_distinct(doy) >= 3
  ) %>%
  group_by(speciesName) %>%
  summarise(
    num_years_condition_satisfied = sum(condition_satisfied)
  )

flowering_data2 <- merge(flowering_data, result_table, by = "speciesName")
flowering_data3 <- flowering_data2 %>% 
filter(!bloom_period == "unknown") %>%  # doesnt contribute to known phenology
  group_by(speciesName) %>%
  mutate(bloomSpecificity = case_when(
    n_distinct(bloom_period) == 1 & bloom_period == 'spring bloomer' ~ "spring bloomer",
    n_distinct(bloom_period) == 1 & bloom_period == 'fall bloomer' ~ "fall bloomer",
     n_distinct(bloom_period) > 1 ~ "opportunistic bloomer"
  )) %>% 
  distinct(speciesName, bloomSpecificity, .keep_all = TRUE)

flowering_data4 <- flowering_data3 %>% 
  filter(num_years_condition_satisfied >= 25)

name_subset <- flowering_data3$speciesName

resolved_subset <- tnrs_match_names(name_subset)
# resolved_subset2 <- filter(resolved_subset, !ott_id == 5502771)
# resolved_subset2 <- filter(resolved_subset2, !ott_id == 173589)
# tree_object <- tol_induced_subtree(ott_ids = ott_id(resolved_subset2)) # appends lineages from open tree

in_tree <- is_in_tree(ott_id(resolved_subset))
in_tree
tr <- tol_induced_subtree(ott_id(resolved_subset)[in_tree])

#ape::plot.phylo(tr, cex = 1.5, type = "phylogram") # edges are being a pain...seems the whole tree was pruned...yay
```

Plotting 
```{r, fig.width=25, fig.height=95}
ott_ids <- tr[[2]]
ott_ids <- gsub("[^0-9 ]", "", ott_ids)
ott_names <- gsub("\\d+", "", tr[[2]])
ott_names <- gsub("\\_ott", "", ott_names)
ott_names <- gsub("_", " ", ott_names)
tree_df <- data.frame(ott_ids, ott_names)

tree_df <- tree_df %>% 
  rename(speciesName = ott_names)

result_df <- tree_df %>%
  left_join(flowering_data3, by = "speciesName")

trait_df <- result_df

#resolved_subset_m <- merge(resolved_tr_subset, flowering_data3, by.x = "unique_name", by.y = "speciesName")

#trait_df <- merge(tree_df, flowering_data3, by.x = "ott_names", by.y = "speciesName",)
tr # try storing the list into this tree objM
# Sort the merged dataframe based on the order of ott_names in tree_df
trait_df <- trait_df[order(match(trait_df$speciesName, tree_df$speciesName)), ]

trait_df <- select(result_df, ott_ids, speciesName, bloomSpecificity, num_years_condition_satisfied)
trait_df <- trait_df %>% 
  mutate(bloomCol = case_when(
    is.na(bloomSpecificity) ~ "grey", 
    bloomSpecificity == "spring bloomer" ~ "darkgreen", 
    bloomSpecificity == "fall bloomer" ~ "darkred", 
    bloomSpecificity == "opportunistic bloomer" ~ "goldenrod"
    
  ))

#ordered_trait_vector <- trait_df$bloomCol[order(match(trait_df$ott_names, trait_df$ott_names))]

tr$traits <- trait_df$bloomCol
png(filename = "/home/jt-miller/Gurlab/sliding-phenology/outputs/flowering-phylogeny.png", width = 10000/2, height = 10000)
ape::plot.phylo(tr, cex = 1.5, type = "phylogram", tip.color = tr$traits) # edges are being a pain...
dev.off()



```
How do the phenometrics look?
```{r, fig.width=25, fig.height=95}
# trait_df <- trait_df %>% 
#   mutate(phenometric_scope =
#            case_when(
#              num_years_condition_satisfied < 10 ~ "grey",
#              num_years_condition_satisfied >= 10 & num_years_condition_satisfied < 20 ~ "steelblue",
#              num_years_condition_satisfied >= 20~ "purple",
#              TRUE ~ "pink"
#              
#            ))
# 
# ape::plot.phylo(tr, cex = 1.5, type = "phylogram", tip.color = trait_df$phenometric_scope) # edges are being a pain...
```


















Some sample data for the package ape 
```{r eval=FALSE, include=FALSE}
data(bird.orders)
### using random colours and thickness
plot(bird.orders,
     edge.color = sample(colors(), length(bird.orders$edge)/2),
     edge.width = sample(1:10, length(bird.orders$edge)/2, replace = TRUE))
title("Random colours and branch thickness")
### rainbow colouring...
X <- c("red", "orange", "yellow", "green", "blue", "purple")
plot(bird.orders,
     edge.color = sample(X, length(bird.orders$edge)/2, replace = TRUE),
     edge.width = sample(1:10, length(bird.orders$edge)/2, replace = TRUE))
title("Rainbow colouring")
plot(bird.orders, type = "c", use.edge.length = FALSE,
     edge.color = sample(X, length(bird.orders$edge)/2, replace = TRUE),
     edge.width = rep(5, length(bird.orders$edge)/2))
segments(rep(0, 6), 6.5:1.5, rep(2, 6), 6.5:1.5, lwd = 5, col = X)
text(rep(2.5, 6), 6.5:1.5, paste(X, "..."), adj = 0)
title("Character mapping...")
plot(bird.orders, "u", font = 1, cex = 0.75)
data(bird.families)
plot(bird.families, "u", lab4ut = "axial", font = 1, cex = 0.5)
plot(bird.families, "r", font = 1, cex = 0.5)
### cladogram with oblique tip labels
plot(bird.orders, "c", FALSE, direction = "u", srt = -40, x.lim = 25.5)
### facing trees with different informations...
tr <- bird.orders
tr$tip.label <- rep("", 23)
layout(matrix(1:2, 1, 2), c(5, 4))
plot(bird.orders, "c", FALSE, adj = 0.5, no.margin = TRUE, label.offset = 0.8,
     edge.color = sample(X, length(bird.orders$edge)/2, replace = TRUE),
     edge.width = rep(5, length(bird.orders$edge)/2))
text(7.5, 23, "Facing trees with\ndifferent informations", font = 2)
plot(tr, "p", direction = "l", no.margin = TRUE,
     edge.width = sample(1:10, length(bird.orders$edge)/2, replace = TRUE))
### Recycling of arguments gives a lot of possibilities
### for tip labels:
plot(bird.orders, tip.col = c(rep("red", 5), rep("blue", 18)),
     font = c(rep(3, 5), rep(2, 17), 1))
plot(bird.orders, tip.col = c("blue", "green"),
     cex = 23:1/23 + .3, font = 1:3)
co <- c(rep("blue", 9), rep("green", 35))
plot(bird.orders, "f", edge.col = co)
plot(bird.orders, edge.col = co)
layout(1)

```

