---
title: "sampling-simus"
author: "JT Miller"
date: "2024-05-08"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Some code designed to simulate the underlying data and how accurately we sample from it under different sampling regimes

Libraries 
```{r}
library(tidyverse)
library(ggplot2)
library(truncnorm)
library(diptest)
```


```{r}
# Using Mike Belitz's method for building a bimodial distribution (see Beltiz et al. 2020)
set.seed(1)
comm_samples <- 1000
sp_samples <- 100
bimodal_comm_doy <- c(rtruncnorm(comm_samples * (2/3), a=0, b=365, mean=150, sd=20),
               rtruncnorm(comm_samples * (1/3), a=0, b=365, mean=220, sd=20))
comm_bimodal_df <- as.data.frame(bimodal_comm_doy)
bimodal_sp_doy <- c(rtruncnorm(sp_samples * (2/3), a=0, b=365, mean=150, sd=10),
               rtruncnorm(sp_samples * (1/3), a=0, b=365, mean=220, sd=10))
sp_bimodal_df <- as.data.frame(bimodal_sp_doy)

ggplot() + 
  geom_histogram(data= comm_bimodal_df, aes(x = bimodal_comm_doy, fill = "Community Level Sample"), bins = 200) + 
  geom_histogram(data= sp_bimodal_df, aes(x = bimodal_sp_doy, fill = "Species Level Sample"), bins = 200) + 
  scale_y_continuous(expand = c(0,0))+
  scale_x_continuous(limits = c(0,365))+
  labs(x = "Day of Year", y = "Number of Individuals") +
  ggtitle("") +
   scale_fill_manual(values = c("Community Level Sample" = "lightgrey", "Species Level Sample" = "black")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(fill = guide_legend(title = NULL))
```
Using Hartigans' Dip Test, test 
H0 = F is a unimodal distribution
H1 = F is at least a bimodal distribution 

Scenario A: Holding all values constant besides sample size 
```{r}
# Constants
my_sd <- 10

# Vector to vary 
n_samples <- c(3:50)
indices <- c(round(n_samples[length(n_samples)/4]), round(n_samples[length(n_samples)/2]), round(n_samples[length(n_samples)/1]))
#selected_values <- n_samples[indices]

# Result vector for storage
dip_results <- rep(NA, length(n_samples))
gs <- list()
# Create a for-loop simulation 
for(i in 1:length(n_samples)){
  bimodal_sp_doy <- c(rtruncnorm(n_samples[i] * (2/3), a=0, b=365, mean=100, sd=my_sd),
                      rtruncnorm(n_samples[i] * (1/3), a=0, b=365, mean=212, sd=my_sd))
  
  # Perform dip test
  dip_test <- dip.test(bimodal_sp_doy, simulate.p.value = TRUE)
  dip_results[i] <- dip_test$p.value
  
  # Make a plotting conditional 
  if(n_samples[i] %in% indices){
    plot_title <- paste("Distribution given", n_samples[i], "\n Samples taken with Standard Deviation =", my_sd)
   g <-ggplot() + 
      geom_histogram(aes(x = bimodal_sp_doy, fill = "Species Level Sample"), bins = 200) + 
      scale_y_continuous(expand = c(0,0))+
      scale_x_continuous(limits = c(0,365))+
      labs(x = "Day of Year", y = "Number of Individuals") +
      ggtitle(plot_title) +
      theme_classic() +
      theme(plot.title = element_text(hjust = 0.5)) +
      guides(fill = guide_legend(title = NULL))
   print(g)

  }
  
}

# how many samples are needed to recover bimodality from the dataset? 
max_n_samples <- NULL
for (i in 1:length(dip_results)) {
  if(all(dip_results[i:length(dip_results)] < 0.05)) {
    max_n_samples <- n_samples[i]
    break
  }
}


# Plot dip test results
ggplot() + 
  geom_point(mapping = aes(x = n_samples, y = dip_results)) +
  geom_line(mapping = aes(x = n_samples, y = 0.05, color = "Critical Value", linetype = "Critical Value")) + 
  geom_line(mapping = aes(x = max_n_samples, y = dip_results, color = "Threshold for Sampling", linetype = "Threshold for Sampling")) +
  scale_color_manual(values = c("Critical Value" = "red", "Threshold for Sampling" = "steelblue")) +
  scale_linetype_manual(values = c("Critical Value" = "dashed", "Threshold for Sampling" = "solid")) +
  theme_bw() + 
  ggtitle("Diptest Probability by Sample Size for Scenario A") +
  xlab("Number of Samples") + 
  ylab("P-Value") +
  theme(
    plot.title = element_text(hjust = 0.5) # Center the title
  ) +
  guides(color = guide_legend(title = "Legend")) # Add legend title

```
Now simulate this 1000 times and take the maximum of the threshold of sampling to determine what number of samples we should use as our baseline. 
```{r}
simu_length <- 100
max_n_samples_storage <- rep(NA, times = simu_length )
for(k in 1:simu_length){
# Constants
my_sd <- 10

# Vector to vary 
n_samples <- c(3:50)
indices <- c(round(n_samples[length(n_samples)/4]), round(n_samples[length(n_samples)/2]), round(n_samples[length(n_samples)/1]))
#selected_values <- n_samples[indices]

# Result vector for storage
dip_results <- rep(NA, length(n_samples))
gs <- list()
# Create a for-loop simulation 
for(i in 1:length(n_samples)){
  bimodal_sp_doy <- c(rtruncnorm(n_samples[i] * (2/3), a=0, b=365, mean=100, sd=my_sd),
                      rtruncnorm(n_samples[i] * (1/3), a=0, b=365, mean=212, sd=my_sd))
  
  # Perform dip test
  dip_test <- dip.test(bimodal_sp_doy, simulate.p.value = TRUE)
  dip_results[i] <- dip_test$p.value
  
  
}

# how many samples are needed to recover bimodality from the dataset? 
max_n_samples <- NULL
for (i in 1:length(dip_results)) {
  if(all(dip_results[i:length(dip_results)] < 0.05)) {
    max_n_samples <- n_samples[i]
    break
  }
}
max_n_samples_storage[k] <- max_n_samples
print(paste("Passed ", k, " iteration"))
}

simu_100_max <- max(max_n_samples_storage) # 27 was max so use that 
```


Varying the means of the two normal distributions simu 
```{r}
# Constants
my_sd <- 10
n_samples <- 27

# Vector to vary 
mu1 <- c(79:152)
mu2 <- c(200:335)

indices <- c(round(mu1[length(mu1)/4]), round(mu1[length(mu1)/2]), round(mu1[length(mu1)/1]))

# Result vectors for storage
dip_results <- matrix(NA, nrow = length(mu1), ncol = length(mu2))
dist_results <- matrix(NA, nrow = length(mu1), ncol = length(mu2))

# Create nested loop simulation 
for(i in 1:length(mu1)){
  for(j in 1:length(mu2)){
    bimodal_sp_doy <- c(rtruncnorm(n_samples* (2/3), a=0, b=365, mean=mu1[i], sd=my_sd),
                        rtruncnorm(n_samples* (1/3), a=0, b=365, mean=mu2[j], sd=my_sd))
    
    # Perform dip test
    dip_test <- dip.test(bimodal_sp_doy, simulate.p.value = TRUE)
    dip_results[i,j] <- dip_test$p.value
    
    # Calculate distance between mu1 and mu2
    dist_results[i,j] <- abs(mu1[i] - mu2[j])
  }
}

# Plot dip test results
library(reshape2)

# Melt the matrices for plotting
dip_results_melted <- melt(dip_results)
dist_results_melted <- melt(dist_results)
colnames(dip_results_melted) <- c("mu1", "mu2", "dip_value")
colnames(dist_results_melted) <- c("mu1", "mu2", "abs_dist")

df_merge <- merge(dip_results_melted, dist_results_melted, by = c("mu1", "mu2"))

df_merge <- df_merge %>% 
  mutate(id = 1:n())
# how many samples are needed to recover bimodality from the dataset? 
df_merge_f <- df_merge %>% 
  filter(dip_value >= 0.05)

max(df_merge_f$id)


max_dist_prior <- subset(df_merge_f, df_merge_f$id == max(df_merge_f$id))
abs_dist_max <- max_dist_prior$abs_dist


# Plot dip test results
ggplot(df_merge) + 
  geom_point(mapping = aes(x = abs_dist , y = dip_value, alpha = 0.1)) +
  geom_line(mapping = aes(x = abs_dist_max, y = dip_value)) +
  theme_bw() + 
  ggtitle("Diptest Probability by Sample Size for Scenario A") +
  xlab("Absolute Distance ") + 
  ylab("P-Value") +
  theme(
    plot.title = element_text(hjust = 0.5) # Center the title
  ) +
  guides(color = guide_legend(title = "Legend")) # Add legend title

```