---
title: "fall-bloomers"
author: "jt-miller"
date: "2024-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fall Bloomers 
The initial sliding window analysis indicated some interesting trends on the shifts in the fall. At the community level, there seems to be a shift towards later flowering periods. This is a relatively unexplored topic in desert phenology and could have significant implications on these taxa's ecology. <br>

Here, I will pull out the taxa that can be sufficiently described as fall blooming taxa by the data and assess species level variation in these phenological trends. 

Load Libraries
```{r}
library(data.table)
library(tidyverse)
```

### Fall subsetting 
Fall is arbitrary in terms of when it actually begins, but there is a distinct period for when early and later (in doy) flowering events occur. Looking at my previous analysis, for plants it appears there is a local minimum ~200 doy. As a first run we will define fall flowering plants as those that bloom >= 200 doy. <br>

The first question then is what taxa (that have sufficient data) are considered spring or fall (or both) blooming plants?
```{r}
hpg <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-plant-data.csv") # load in high coverage flowering plants


```



