---
title: "pulling-out-phenology"
author: "jt-miller"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Taking the bee and plant data and pulling out the relevant phenometerics by quantile using 
1. KDE1d distributions
2. PHENESSE 

Load Libraries
```{r}
library(kde1d)
library(dplyr)
library(ggplot2)
library(phenesse)
library(data.table)
library(stringr)
library(foreach)
library(doParallel)
library(ggpubr)
```

Load Data
```{r}
hbg <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-bee-data.csv")
hpg <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-plant-data.csv")
```

Create datasets according to geographic bins 
```{r}
# bee taxa
grid_holder_b <- list()
gid_b_v <- unique(hbg$GID)
for(i in 1:length(unique(hbg$GID))){
  grid_holder_b[[i]] <- hbg[GID == gid_b_v[[i]],]
}
# plant taxa
grid_holder_p <- list()
gid_p_v <- unique(hpg$GID)
for(i in 1:length(unique(hbg$GID))){
  grid_holder_p[[i]] <- hpg[GID == gid_p_v[[i]],]
}
```


Create Sliding Decadal bins of datasets
```{r}
# Function that uses a for-loop to create the decadal subsets
decadal_subsetter <- function(dataset, start_year, end_year, bin_size){
  decadal_subsets <- list() # empty storage list
for(i in start_year:(end_year - bin_size)){
  end <- i + bin_size # Our end condition 
  subset_data <- dataset[year >= i & year <= end, ] # Subsets the data into bins on a decadal sliding scale
  decadal_subsets[[paste(i, end, sep = "-")]] <- subset_data # Label
}
  return(decadal_subsets)
}

# Set arguments
start_year <- 1950
end_year <- 2023
bin_size <- 10
gids <- c(217, 229, 230, 241)

nested_list_holder_b <- list()
for(i in 1:length(grid_holder_b)){
decadal_subsets <- decadal_subsetter(dataset = grid_holder_b[[i]], # Take the ith grid and do the decadal subsetting  
                  start_year = start_year, 
                  end_year = end_year, 
                  bin_size = bin_size)
nested_list_holder_b[[paste(gids[i], "grid", sep = "-")]] <- decadal_subsets
}

nested_list_holder_p <- list()
for(i in 1:length(grid_holder_p)){
decadal_subsets <- decadal_subsetter(dataset = grid_holder_p[[i]], # Take the ith grid and do the decadal subsetting  
                  start_year = start_year, 
                  end_year = end_year, 
                  bin_size = bin_size)
nested_list_holder_p[[paste(gids[i], "grid", sep = "-")]] <- decadal_subsets
}


```


Creating Table Visualizations
```{r}
library(knitr)
sample_bin <- nested_list_holder_p[[1]]
sample_df <- data.frame()
for(i in 1:length(sample_bin)){
  decadal_bin_name <- names(sample_bin[i])
  number_of_records <- nrow(sample_bin[[i]])
  number_of_species <- length(unique(sample_bin[[i]]$scientificName))
  
  current_result <- data.frame(
    DecadalBinName = decadal_bin_name,
    NumberOfRecords = number_of_records,
    NumberOfSpecies = number_of_species)
    
    # Add the data frame to the growing dataframe
  sample_df <- rbind(sample_df, current_result)
  
}
header <- head(sample_df, 5)
tailer <- tail(sample_df, 5)

samp <- rbind(header, tailer)

ggtexttable(samp, cols = c("Decadal Bin", "Number of Records", "Number of Species")) 
ggsave("/home/jt-miller/Gurlab/sliding-phenology/outputs/graphical-outputs/decadal-bin-table.png")

```

### Species Level kernel density estimations
```{r}
for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[w]) # replace with w
  spatial_bin <- nested_list_holder_b[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    for(i in 1:length(unique_names_in_bin_v)){
      named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(named_subset) <= 5 | length(unique(named_subset$doy)) <= 3){ # adjust to 3 days***
    named_phenometric_summary <- data.table(taxa = "bee", scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    named_subset <- named_subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = named_subset$doy) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_plots/", "bee_", name, "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0(unique(named_subset$scientificName), " kernel density distribution for time bin ", spatial_temporal_ID), xlab = "day of year", xlim = c(0,365))
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(taxa = "bee", scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", "bee_", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", "bee_", name, "_", spatial_temporal_ID, ".csv"))
    } # end of i loop
  } # end of j loop
} # end of w loop

#### And for Plants ################################################################################################################


for(w in 1:length(nested_list_holder_p)){
  spatial_ID <- names(nested_list_holder_p[w]) # replace with w
  spatial_bin <- nested_list_holder_p[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    for(i in 1:length(unique_names_in_bin_v)){
      named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(named_subset) <= 5 | length(unique(named_subset$doy)) <= 3){ # adjust to 3 days***
    named_phenometric_summary <- data.table(taxa = "plant", scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    named_subset <- named_subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = named_subset$doy) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_plots/", "plant_", name, "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0(unique(named_subset$scientificName), " kernel density distribution for time bin ", spatial_temporal_ID), xlab = "day of year", xlim = c(0,365))
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(taxa = "plant", scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", "plant_", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", "plant_", name, "_", spatial_temporal_ID, ".csv"))
    } # end of i loop
  } # end of j loop
} # end of w loop
```

### Community Level Kernel Desnity Estimations
```{r}
# Bees
for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[w]) # replace with w
  spatial_bin <- nested_list_holder_b[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
    #for(i in 1:length(unique_names_in_bin_v)){
      #named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(subset) <= 5 | length(unique(subset$doy)) <= 3){
    named_phenometric_summary <- data.table(taxa = "bees", GID = unique(subset$GID), number_of_records = nrow(subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_plots/", "bees", "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0("Bee Community ", " Kernel Density Distribution for Time Bin ", spatial_temporal_ID), xlab = "day of year")
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(taxa = "bees", GID = unique(subset$GID),
                                            number_of_records = nrow(subset), time_bin = named_temporal_bin)
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
    named_phenometric_summary <- named_phenometric_summary[, num_of_species := num_of_species]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_tables/", "bees", "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_tables/", "bees", "_", spatial_temporal_ID, ".csv"))
    #} # end of i loop
  } # end of j loop
} # end of w loop

# Plants 
for(w in 1:length(nested_list_holder_p)){
  spatial_ID <- names(nested_list_holder_p[w]) # replace with w
  spatial_bin <- nested_list_holder_p[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
    #for(i in 1:length(unique_names_in_bin_v)){
      #named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(subset) <= 5 | length(unique(subset$doy)) <= 3){
    named_phenometric_summary <- data.table(taxa = "plants", GID = unique(subset$GID), number_of_records = nrow(subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density. xmin = 0, xmax = 365
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_plots/", "plants", "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0("Plant Community ", " Kernel Density Distribution for Time Bin ", spatial_temporal_ID), xlab = "day of year", xlim = c(0,365))
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(taxa = "plants", GID = unique(subset$GID),
                                            number_of_records = nrow(subset), time_bin = named_temporal_bin)
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
    named_phenometric_summary <- named_phenometric_summary[, num_of_species := num_of_species]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_tables/", "plants", "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_tables/", "plants", "_", spatial_temporal_ID, ".csv"))
    #} # end of i loop
  } # end of j loop
} # end of w loop
```

### Extract Community Level Trends 
```{r}
directory <- "/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_community_tables/"
file_list <- list.files(directory, full.names = TRUE)

community_phenometrics <- list()
for(i in 1:length(file_list)){
  community_phenometrics[[i]] <- fread(file_list[[i]])

}
community_phenometrics <- rbindlist(community_phenometrics)

community_phenometrics$time_bin <- factor(community_phenometrics$time_bin, levels = unique(community_phenometrics$time_bin))

plot(community_phenometrics, x = community_phenometrics$time_bin, y = community_phenometrics$fiftieth_percentile)

plant_comm_metrics <- community_phenometrics[taxa == "plants"]
ggplot(plant_comm_metrics, aes(x = time_bin, y = fiftieth_percentile)) +
  geom_point() +
  labs(title = "Temporal Plot of Medians",
       x = "Time Bins",
       y = "Medians") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~GID)

bees_comm_metrics <- community_phenometrics[taxa == "bees"]
ggplot(bees_comm_metrics, aes(x = time_bin, y = fiftieth_percentile)) +
  geom_point() +
  labs(title = "Temporal Plot of Medians",
       x = "Time Bins",
       y = "Medians") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~GID)


```
### Community level trends showing example of distributed data -> kernel density map
```{r}
example <- nested_list_holder_p[[4]][[23]]
grid <- names(nested_list_holder_p)[[4]]
time_bin <- names(nested_list_holder_p[[4]][23])
grid <- gsub("-", " ", grid)
# Split the string into words
words <- strsplit(grid, " ")[[1]]
# Reorder the words
grid <- paste(words[2], words[1], collapse = " ")
spatial_temporal_ID <- paste0(time_bin, " ", grid)

png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/graphical-outputs/", "histogram-plants", "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
hist(example$doy, main = paste0("Plant Community Flowering Records for Time Bin", " ", spatial_temporal_ID), xlab = "Day of Year")
dev.off
subset <- example[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density. xmin = 0, xmax = 365
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/graphical-outputs/", "plants-kernel-sample", "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0("Plant Community", " Kernel Density Distribution for Time Bin ", spatial_temporal_ID), xlab = "day of year", xlim = c(0,365))
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 

```

### Community level trends on the same plots (Plants)
```{r}

# Create an empty storage df  
result_df <- data.frame()
nested_kde_holder <- list()

for(w in 1:length(nested_list_holder_p)){
  spatial_ID <- names(nested_list_holder_p[w]) # replace with w
  spatial_bin <- nested_list_holder_p[[w]] # replace with w 
  kde_holder <- list() # intialize a ist to store kde sims for each w
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
       if(nrow(subset) <= 5 | length(unique(subset$doy)) <= 3){
         print("Skipping Plot Due to Insufficient Data")
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density. xmin = 0, xmax = 365
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Store data 
    row_data <- data.frame(spatialTemporalID = spatial_temporal_ID, numberSpecies = num_of_species, quant50=quant50, var=var50)
    result_df <- rbind(result_df, row_data)
    # Store kde.sims in nested structure
      kde_holder[[named_temporal_bin]] <- list(
        quant50 = quant50,
        var50 = var50,
        kde.simulations = kde.sims)
    

} # End of else statement
    
  } # end of j loop
  
  nested_kde_holder[[spatial_ID]] <- kde_holder # Store the kde_holder for the current 'w' in nested_kde_holder
} # end of w loop

# Modified code to include legends with years ########################################################################################################
library(RColorBrewer)
png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/graphical-outputs/", "grid-cell-241-sample-dists", ".png"), width = 800, height = 600)
plot(1, type = "n", main = "Plant Community Kernel Density Distribution by Decadal Bin for Grid Cell 241", xlab = "Decadal Bin", ylab = "Density", xlim = c(0, 400), ylim = c(0, 0.015))

# Extract the first 5 and last 5 time bins
cell_241 <- nested_kde_holder[[4]]
time_bins_to_plot <- c(cell_241[1:5], cell_241[(length(cell_241)-4):length(cell_241)])

# Use a color palette for lines
#line_colors <- c("lightblue", "lightgreen", "lightorange", "lightpurple", "lightred", "darkblue", "darkgreen", "darkorange", "darkpurple", "darkred")

light_colors <- brewer.pal(5, "Set3")
dark_colors <- brewer.pal(5, "Dark2")
line_colors <- c(light_colors, dark_colors)
# Create a vector to store legend labels
years <- names(time_bins_to_plot)

# Now plot these kernel densities iteratively 
for (w in 1:length(time_bins_to_plot)) {
  spatial_id <- names(time_bins_to_plot[w]) # replace with w
  spatial_bin <- time_bins_to_plot[[w]] # replace with w
  year <-  years[w] # Assuming the year information is stored in the first element

  legend_labels[w] <- paste("Kernel Density (", year, ")", sep = "")

  for (j in 1:length(spatial_bin)) {
    quant50 <- spatial_bin[1][[1]]
    kde_sim <- spatial_bin[[3]]
    var50 <- spatial_bin[[2]]
    
    # Plot kernel density using lines with different colors
    lines(kde_sim, col = line_colors[w], xaxt = "n")

    # Add vertical line for 50th percentile
    abline(v = quant50, col = line_colors[w], lty = 2)

    # Add error bars
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    segments(x0 = quant50, y0 = lower_bound, x1 = quant50, y1 = upper_bound, col = line_colors[w], lwd = 2)
  }
} # end of w loop.

# Add legend outside the loop to avoid duplication
legend("topright", legend = legend_labels[1:10], col = line_colors[1:length(time_bins_to_plot)], lty = 1, lwd = 1, cex = 0.75)
dev.off()
```
### Community level trends on the same plots (Bees)
```{r}
# Create an empty storage df  
result_df <- data.frame()
nested_kde_holder <- list()

for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[w]) # replace with w
  spatial_bin <- nested_list_holder_b[[w]] # replace with w 
  kde_holder <- list() # intialize a ist to store kde sims for each w
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
       if(nrow(subset) <= 5 | length(unique(subset$doy)) <= 3){
         print("Skipping Plot Due to Insufficient Data")
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy) # est the univariate likelihood kernel density. xmin = 0, xmax = 365
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Store data 
    row_data <- data.frame(spatialTemporalID = spatial_temporal_ID, numberSpecies = num_of_species, quant50=quant50, var=var50)
    result_df <- rbind(result_df, row_data)
    # Store kde.sims in nested structure
      kde_holder[[named_temporal_bin]] <- list(
        quant50 = quant50,
        var50 = var50,
        kde.simulations = kde.sims)
    

} # End of else statement
    
  } # end of j loop
  
  nested_kde_holder[[spatial_ID]] <- kde_holder # Store the kde_holder for the current 'w' in nested_kde_holder
} # end of w loop

png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/graphical-outputs/", "grid-cell-241-sample-dists-bees", ".png"), width = 800, height = 600)
plot(1, type = "n", main = "Bee Community Kernel Density Distribution by Decadal Bin for Grid Cell 241", xlab = "Decadal Bin", ylab = "Density", xlim = c(0, 400), ylim = c(0, 0.05))

# Extract the first 5 and last 5 time bins
cell_241 <- nested_kde_holder[[4]]
time_bins_to_plot <- c(cell_241[1:5], cell_241[(length(cell_241)-4):length(cell_241)])

# Use a color palette for lines
#line_colors <- c("lightblue", "lightgreen", "lightorange", "lightpurple", "lightred", "darkblue", "darkgreen", "darkorange", "darkpurple", "darkred")

light_colors <- brewer.pal(5, "Set3")
dark_colors <- brewer.pal(5, "Dark2")
line_colors <- c(light_colors, dark_colors)
# Create a vector to store legend labels
years <- names(time_bins_to_plot)

# Now plot these kernel densities iteratively 
for (w in 1:length(time_bins_to_plot)) {
  spatial_id <- names(time_bins_to_plot[w]) # replace with w
  spatial_bin <- time_bins_to_plot[[w]] # replace with w
  year <-  years[w] # Assuming the year information is stored in the first element

  legend_labels[w] <- paste("Kernel Density (", year, ")", sep = "")

  for (j in 1:length(spatial_bin)) {
    quant50 <- spatial_bin[1][[1]]
    kde_sim <- spatial_bin[[3]]
    var50 <- spatial_bin[[2]]
    
    # Plot kernel density using lines with different colors
    lines(kde_sim, col = line_colors[w], xaxt = "n")

    # Add vertical line for 50th percentile
    abline(v = quant50, col = line_colors[w], lty = 2)

    # Add error bars
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    segments(x0 = quant50, y0 = lower_bound, x1 = quant50, y1 = upper_bound, col = line_colors[w], lwd = 2)
  }
} # end of w loop.

# Add legend outside the loop to avoid duplication
legend("topright", legend = legend_labels[1:10], col = line_colors[1:length(time_bins_to_plot)], lty = 1, lwd = 1, cex = 0.75)
dev.off()
```

### Species level phenology trends
```{r}
directory <- "/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/"
file_list <- list.files(directory, full.names = TRUE)

species_lvl_phenometrics <- list()
for(i in 1:length(file_list)){
  holding_dt <- fread(file_list[[i]])
  pattern <-  "(\\d{4}-\\d{4})" # Create a pattern for the date
  date <- str_extract(file_list[[i]], pattern)
  holding_dt <- holding_dt[, time_bin := date]
species_lvl_phenometrics[[i]] <- holding_dt
}
species_lvl_phenometrics <- rbindlist(species_lvl_phenometrics)

plant_species_lvl_phenometrics <- species_lvl_phenometrics[taxa == "plant"]
plant_species_lvl_phenometrics <- plant_species_lvl_phenometrics[!is.na(fiftieth_percentile)]
bee_species_lvl_phenometrics <- species_lvl_phenometrics[taxa == "plant"]
bee_species_lvl_phenometrics <- bee_species_lvl_phenometrics[!is.na(fiftieth_percentile)]

# Create a custom order for the time_bin factor
factor_variable_ordered <- plant_species_lvl_phenometrics$time_bin[order(plant_species_lvl_phenometrics$time_bin)]
factor_variable_ordered <- unique(factor_variable_ordered)
print(unique(plant_species_lvl_phenometrics$time_bin))
# Reorder the time_bin factor levels
plant_species_lvl_phenometrics$time_bin <- factor(plant_species_lvl_phenometrics$time_bin, levels = factor_variable_ordered)

# Create boxplots using ggplot
ggplot(data = plant_species_lvl_phenometrics, mapping = aes(x = time_bin, y = fiftieth_percentile, fill = time_bin)) +
  geom_boxplot() +
  labs(title = "Plants Species Level Median Boxplots",
       x = "Time Bins",
       y = "Medians") + 
  guides(fill = FALSE) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~GID)


# Create a custom order for the time_bin factor
factor_variable_ordered <- bee_species_lvl_phenometrics$time_bin[order(bee_species_lvl_phenometrics$time_bin)]
factor_variable_ordered <- unique(factor_variable_ordered)
print(unique(bee_species_lvl_phenometrics$time_bin))
# Reorder the time_bin factor levels
bee_species_lvl_phenometrics$time_bin <- factor(bee_species_lvl_phenometrics$time_bin, levels = factor_variable_ordered)

ggplot(data = bee_species_lvl_phenometrics, mapping = aes(x = time_bin, y = fiftieth_percentile)) +
  geom_boxplot() +
  labs(title = "Plants Species Level Median Boxplots",
       x = "Time Bins",
       y = "Medians") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~GID)

```
### The following code is an outline for Phenesse, DNR here needs parallel computing to finish within a reasonable timeframe: see sliding-window-phenesse.R and its associated SLURM script for Run Details. 
PHENESSE
```{r eval=FALSE, include=FALSE}
### And just the weib_percentile function from phenesse
for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[w]) # replace with w
  spatial_bin <- nested_list_holder_b[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    for(i in 1:length(unique_names_in_bin_v)){
      named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(named_subset) <= 5 | length(unique(named_subset$doy)) <= 3){
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using phenesse
    named_subset <- named_subset[doy >= 0 & doy <= 365]
     fiftieth_est <- tryCatch(phenesse::weib_percentile(observations = named_subset$doy, percentile = 0.5, iterations = 500),
                                error = function(e) NA)
     # Start plot saving
     name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
     name <- gsub(" ", "-", name) # ""
     png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/phenesse_weib_plots/", name, "_", spatial_temporal_ID, ".png"), width = 1000, height = 600)
     hist(named_subset$doy, xlim = c(0,365), main = paste0(unique(named_subset$scientificName), " histogram for time bin ", spatial_temporal_ID), xlab = "day of year")
     abline(v = fiftieth_est, col = "red")
     legend("topright", legend ="50th Percentile Estimate",
       col = "red", lty =  1, lwd = 1, cex = 0.9)
     dev.off()
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := fiftieth_est]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/phenesse_weib_tables/", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/phenesse_weib_tables/", name, "_", spatial_temporal_ID, ".csv"))
    } # end of i loop
  } # end of j loop
} # end of w loop

```



Kernel Density xmin xmax issue (see notes)
```{r eval=FALSE, include=FALSE}
# Checking some oddities 
spatial_bin <- nested_list_holder_p[[1]] # Grab the first GID
test <- spatial_bin[[23]] # Grab 1972-1982
test <- test[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = test$doy) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
plot(kde.sims, main = paste0("Plant Community ", " Kernel Density Distribution for Time Bin ", spatial_temporal_ID), xlab = "day of year")
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)

hist(x =test$doy)
####
  kde.sims <- kde1d(x = test$doy, xmin = 0) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance

plot(kde.sims, main = paste0("Plant Community ", " Kernel Density Distribution for Time Bin ", spatial_temporal_ID), xlab = "day of year")
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)

hist(x =test$doy)
   
```