---
title: "time-binning"
author: "jt-miller"
date: "2023-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Assessing the Data's temporal distribution in order to inform phenology analysis 
Load Libraries
```{r}
library(data.table)
library(lubridate)
library(ggplot2)
library(dplyr)
```
Bring in the Sonoran Flowering Data & Bee Data (cleaned in data-cleaning.Rmd)
```{r}
sonoran_flowering_data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/flowering_sonoran_data.csv")
cch2_fl_sonoran_data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/cch2_fl_sonoran_data.csv")
sonoran_bee_data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/bee_occ_sonoran_data.csv")
```

Merge Flowering Datasets 
```{r}
# Add a field denoting aggregator
sfd <- sonoran_flowering_data[, aggregator := "inat"] 
cch2 <- cch2_fl_sonoran_data[, aggregator := "cch2"]

# Assure correct date class in the cch2 data
cch2$eventDate <- as.POSIXct(cch2$eventDate, format = "%Y-%m-%d")

fl_data <- bind_rows(sfd, cch2) # dplyr function that rbinds the data, and replaces NAs for fields missing from eachothers dataset 

bd <- sonoran_bee_data # rename for simplicity 
```


Parse out year month day and day of year (doy) fields
```{r}
# Create new temporal fields for flowering plants 
fld <- fl_data[, year := year(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fl_data[, month := month(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fl_data[, day := day(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fl_data[, doy := yday(as.POSIXlt.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]

# Do the same thing for the bees
bd$eventDate <- as.POSIXct(bd$eventDate, format = "%Y/%m/%d %H:%M:%S", tz = "UTC")
bd <- bd[, c("year", "month", "day", "doy") := list(year(eventDate), month(eventDate), day(eventDate), yday(eventDate))] # Shortcut

fld[, (.N), by = "year"] 
bd[, (.N), by = "year"] 
class(fld$year)
class(bd$year)
fld <- fld[year >= 1950]
bd <- bd[year >= 1950]
```

First, phenology density mapping over full time scale 


55 changes: 55 additions & 0 deletions 55
R/rmds/data-cleaning.Rmd
@@ -152,3 +152,58 @@ class(cch2_fl_sonoran_data)
fwrite(cch2_fl_sonoran_data, "/home/jt-miller/Gurlab/sliding-phenology/data/processed/cch2_fl_sonoran_data.csv")
```
### Now clean the bee data 

Bring in the cleaned bee dataset produced in Dorey et al. 2023
```{r}
cleaned_bees_occur <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/raw/Bee-Cleaned-Data/OutputData/05_cleaned_database.csv")
```

Check data for flags
```{r}
names(cleaned_bees_occur)
```
Usual cleaning for my research questions + make it spatial
```{r}
# Find fields of interest. 
names(cleaned_bees_occur)
bee_occ <- cleaned_bees_occur[, .(id, informationWithheld, recordedBy, identifiedBy, eventDate, verbatimEventDate, decimalLatitude, decimalLongitude, coordinateUncertaintyInMeters, scientificName, taxonRank)]
# Remove data without eventDate
bee_occ <- bee_occ[!is.na(eventDate)]
# Remove data without lon lat 
bee_occ <- bee_occ[!is.na(decimalLatitude)]
bee_occ <- bee_occ[!is.na(decimalLongitude)]
# Make our inat data spatial 
bee_occ_sf <- st_as_sf(bee_occ,
                              coords = c("decimalLongitude", "decimalLatitude"),
                              crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs",
                              remove = FALSE)
# Bring in Ecoregion shapefiles 
na_ecoregions_3 <- sf::read_sf("/home/jt-miller/Gurlab/sliding-phenology/data/raw/shapefiles/NA-level-3-epa-ecoregions/NA_CEC_Eco_Level3.shp")
# Filter for just the Sonoran Desert & Mojave
sonoran_shp <- subset(na_ecoregions_3, na_ecoregions_3$NA_L3NAME == "Sonoran Desert")
# Convert the crs of the inat data to that of the sonoran shapefile
bee_occ_sf <- bee_occ_sf %>% 
  st_transform(crs = st_crs(sonoran_shp))
# Trim
bee_occ_sonoran_data <- bee_occ_sf[sonoran_shp, ]
```

Write the processed bee records out to a file to work with later
```{r}
# Drop spatial status 
bee_occ_sonoran_data <- bee_occ_sonoran_data %>% 
  st_drop_geometry()
class(bee_occ_sonoran_data)
fwrite(bee_occ_sonoran_data, "/home/jt-miller/Gurlab/sliding-phenology/data/processed/bee_occ_sonoran_data.csv")
```
57 changes: 43 additions & 14 deletions 57
R/rmds/time-binning.Rmd
@@ -16,54 +16,83 @@ library(lubridate)
library(ggplot2)
library(dplyr)
```
Bring in the Sonoran Flowering Data (cleaned in data-cleaning.Rmd)
Bring in the Sonoran Flowering Data & Bee Data (cleaned in data-cleaning.Rmd)
```{r}
sonoran_flowering_data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/flowering_sonoran_data.csv")
cch2_fl_sonoran_data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/cch2_fl_sonoran_data.csv")
sonoran_bee_data <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/bee_occ_sonoran_data.csv")
```

Merge Datasets 
Merge Flowering Datasets 
```{r}
# Add a field denoting aggregator
sfd <- sonoran_flowering_data[, aggregator := "inat"] 
cch2 <- cch2_fl_sonoran_data[, aggregator := "cch2"]
# Assure correct date class in the cch2 data
cch2$eventDate <- as.Date(cch2$eventDate)
cch2$eventDate <- as.POSIXct(cch2$eventDate, format = "%Y-%m-%d")
#cch2$eventDate <- as.Date(cch2$eventDate)
fl_data <- bind_rows(sfd, cch2) # dplyr function that rbinds the data, and replaces NAs for fields missing from eachothers dataset
fl_data <- bind_rows(sfd, cch2) # dplyr function that rbinds the data, and replaces NAs for fields missing from eachothers dataset 
bd <- sonoran_bee_data # rename for simplicity 
```


Parse out year month day and day of year (doy) fields
```{r}
fld <- fl_data[, year := year(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fl_data[, month := month(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fl_data[, day := day(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fl_data[, doy := yday(as.POSIXlt.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
# Create new temporal fields for flowering plants 
fld <- fl_data[, year := year(as.POSIXct(eventDate, format = "%Y-%m-%d"))]
fld <- fld[, month := month(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fld[, day := day(as.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
fld <- fld[, doy := yday(as.POSIXlt.POSIXct(eventDate, format = "%Y-%m-%d %H:%M:%S %Z"))]
# Do the same thing for the bees
bd$eventDate <- as.POSIXct(bd$eventDate, format = "%Y/%m/%d %H:%M:%S", tz = "UTC")
bd <- bd[, c("year", "month", "day", "doy") := list(year(eventDate), month(eventDate), day(eventDate), yday(eventDate))] # Shortcut
fld[, (.N), by = "year"] # Looks like very few records prior to the 2000s, ask Rob but as an arbitrary cut off lets go from there
fld[, (.N), by = "year"] 
bd[, (.N), by = "year"] 
class(fld$year)
class(bd$year)
fld <- fld[year >= 1950]
bd <- bd[year >= 1950]
```

First, phenology density mapping over full time scale 
```{r}
### For Plants 
ggplot(fld, aes(x = year)) +
  geom_density(fill = "darkgreen", alpha = 0.5) +
  labs(title = "Density Time Series of Flowering Observations per Year",
       x = "Year",
       y = "Density") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(fld, aes(x = factor(year))) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Number of Flowering Observations per Year",
       x = "Year",
       y = "Count (N)") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

### For Bees 
ggplot(bd, aes(x=year)) + 
  geom_density(fill = "darkorange", alpha = 0.5) + 
  labs(title = "Density Time Series of Bee Observations per Year",
       x = "Year", 
       y = "Density") + 
  theme_minimal()
ggplot(bd, aes(x = factor(year))) +
  geom_bar(fill = "darkorange") +
  labs(title = "Number of Bee Observations per Year",
       x = "Year",
       y = "Count (N)") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```






