---
title: "pulling-out-phenology"
author: "jt-miller"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Taking the bee and plant data and pulling out the relevant phenometerics by quantile using 
1. KDE1d distributions
2. PHENESSE 

Load Libraries
```{r}
library(kde1d)
library(dplyr)
library(ggplot2)
library(phenesse)
library(data.table)
```

Load Data
```{r}
hbg <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-bee-data.csv")
hpg <- fread("/home/jt-miller/Gurlab/sliding-phenology/data/processed/high-coverage-plant-data.csv")
```

Create datasets according to geographic bins 
```{r}
grid_holder_b <- list()
gid_b_v <- unique(hbg$GID)
for(i in 1:length(unique(hbg$GID))){
  grid_holder_b[[i]] <- hbg[GID == gid_b_v[[i]],]
}
```


Create Sliding Decadal bins of datasets
```{r}
# Function that uses a for-loop to create the decadal subsets
decadal_subsetter <- function(dataset, start_year, end_year, bin_size){
  decadal_subsets <- list() # empty storage list
for(i in start_year:(end_year - bin_size)){
  end <- i + bin_size # Our end condition 
  subset_data <- dataset[year >= i & year <= end, ] # Subsets the data into bins on a decadal sliding scale
  decadal_subsets[[paste(i, end, sep = "-")]] <- subset_data # Label
}
  return(decadal_subsets)
}

# Set arguments
start_year <- 1950
end_year <- 2023
bin_size <- 10
nested_list_holder_b <- list()
for(i in 1:length(grid_holder_b)){
decadal_subsets <- decadal_subsetter(dataset = grid_holder_b[[i]], # Take the ith grid and do the decadal subsetting  
                  start_year = start_year, 
                  end_year = end_year, 
                  bin_size = bin_size)
nested_list_holder_b[[paste(i, "grid", sep = "-")]] <- decadal_subsets
}
```


For each decadal bin, pull out each species and evaluate the phenometrics. Create a df for each species per decadal bin that has the phenometric quantile 50%. 
*Note if there is insufficient data, NA should populate the field 
```{r}
for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[w]) # replace with w
  spatial_bin <- nested_list_holder_b[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    for(i in 1:length(unique_names_in_bin_v)){
      named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(named_subset) <= 7 | length(unique(named_subset$doy)) <= 2){
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    named_subset <- named_subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = named_subset$doy, xmin = 0, xmax = 365) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_plots/", name, "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0(unique(named_subset$scientificName), " kernel density distribution for time bin ", spatial_temporal_ID), xlab = "day of year")
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", name, "_", spatial_temporal_ID, ".csv"))
    } # end of i loop
  } # end of j loop
} # end of w loop

### And just the weib_percentile function from phenesse
for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[w]) # replace with w
  spatial_bin <- nested_list_holder_b[[w]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[j]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[j]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    for(i in 1:length(unique_names_in_bin_v)){
      named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(named_subset) <= 7 | length(unique(named_subset$doy)) <= 2){
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using phenesse
    named_subset <- named_subset[doy >= 0 & doy <= 365]
     fiftieth_est <- tryCatch(phenesse::weib_percentile(observations = named_subset$doy, percentile = 0.5, iterations = 500),
                                error = function(e) NA)
     # Start plot saving
     name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
     name <- gsub(" ", "-", name) # ""
     png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/phenesse_weib_plots/", name, "_", spatial_temporal_ID, ".png"), width = 1000, height = 600)
     hist(named_subset$doy, xlim = c(0,365), main = paste0(unique(named_subset$scientificName), " histogram for time bin ", spatial_temporal_ID), xlab = "day of year")
     abline(v = fiftieth_est, col = "red")
     legend("topright", legend ="50th Percentile Estimate",
       col = "red", lty =  1, lwd = 1, cex = 0.9)
     dev.off()
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := fiftieth_est]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/phenesse_weib_tables/", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/phenesse_weib_tables/", name, "_", spatial_temporal_ID, ".csv"))
    } # end of i loop
  } # end of j loop
} # end of w loop

```

Lets also try a community wide phenometric distribution while we're at it.
```{r}
for(w in 1:length(nested_list_holder_b)){
  spatial_ID <- names(nested_list_holder_b[1]) # replace with w
  spatial_bin <- nested_list_holder_b[[1]] # replace with w 
  for(j in 1:length(spatial_bin)){
    named_temporal_bin <- names(spatial_bin[1]) # replace with j
    spatial_temporal_ID <- paste0(spatial_ID, "_", named_temporal_bin)
    temporal_spatial_bin <- spatial_bin[[1]] # replace with j
    unique_names_in_bin_v <- unique(temporal_spatial_bin$scientificName) # extract a vector of uniq scientific names from bin
    num_of_species <- length(unique_names_in_bin_v)
    subset <- temporal_spatial_bin # rename
    #for(i in 1:length(unique_names_in_bin_v)){
      #named_subset <- temporal_spatial_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
       if(nrow(subset) <= 7 | length(unique(subset$doy)) <= 2){
    named_phenometric_summary <- data.table(taxa = "Bees", GID = unique(subset$GID), number_of_records = nrow(subset), fiftieth_percentile = NA, fiftieth_var = NA, upper_var_bound = NA, lower_var_bound = NA)
  }else{
    # Using kde1
    subset <- subset[doy >= 0 & doy <= 365]
    kde.sims <- kde1d(x = subset$doy, xmin = 0, xmax = 365) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    name <- gsub(",", "", unique(subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_plots/", name, "_", spatial_temporal_ID, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0(unique(subset$scientificName), " kernel density distribution for time bin ", spatial_temporal_ID), xlab = "day of year")
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(scientificName = unique(subset$scientificName), GID = unique(subset$GID),
                                            number_of_records = nrow(subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
  
fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", name, "_", spatial_temporal_ID, ".csv"))

} # End of else statement
    name <- gsub(",", "", unique(subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    fwrite(named_phenometric_summary, paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_tables/", name, "_", spatial_temporal_ID, ".csv"))
    #} # end of i loop
  } # end of j loop
} # end of w loop
```


```{r}
# Initialize objects outside loops
result_list <- list()

for (j in 1:length(nested_grid_holder_b)) {
  
  spatial_bin <- nested_list_holder_b[1] # replace with j
  name_of_spatial_bin <- names(spatial_bin)
  
  for (w in 1:length(spatial_bin[[1]])) {
    
    spatial_times_bin <- spatial_bin[[1]] # replace with w

    for (i in 1:length(unique_names_in_bin_v)) {
      
      spatial_time_subset <- spatial_time_bin[[1]] # replace with i
      unique_names_in_bin <- unique(spatial_time_subset$scientificName)
      spatial_time_subset_name <- names(spatial_time_subset)
      
      named_subset <- test_bin[scientificName %in% unique_names_in_bin_v[i]] # replace with i
  if(nrow(named_subset) <= 7){
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fifieth_percentile = NA)
  }else{
    # Using kde1
    kde.sims <- kde1d(x = named_subset$doy, xmin = 0, xmax = 365) # est the univariate likelihood kernel density
    quant50 <- qkde1d(p = 0.50, obj=kde.sims) # pull out the 50th quantile
    emp.rand.samp <- rkde1d(n = 10000, obj = kde.sims) # use random generation to find variance
    var50 <- var(emp.rand.samp) # calc variance
    
    # Plot + Variance #############################################################################################################
    name <- gsub(",", "", unique(named_subset$scientificName)) # Fix some file naming shennanigans
    name <- gsub(" ", "-", name) # ""
    png(paste0("/home/jt-miller/Gurlab/sliding-phenology/outputs/kde_plots/", name, "_", test_bin_year, ".png"), width = 800, height = 600)
    plot(kde.sims, main = paste0(unique(named_subset$scientificName), " kernel density distribution for time bin ", test_bin_year), xlab = "day of year")
    abline(v = quant50, col = "red")
    upper_bound <- quant50 + sqrt(var50)
    lower_bound <- quant50 - sqrt(var50)
    # Add error bars
    #segments(x0 = quant50, y0 = lower_boun, x1 = quant50, y1 = upper_bound, col = "blue", lwd = 2)
    abline(v = upper_bound, lty= 'dashed', col = "blue")
    abline(v = lower_bound, lty = 'dashed', col = "blue")
    # Add a legend
    legend("topright", legend = c("Kernel Density", "50th Percentile", "Variance"),
       col = c("black", "red", "blue"), lty = c(1, 1, 2), lwd = c(1, 1, 2), cex = 0.9)
   
     dev.off() 
     # End plot saving ##################################################################################################
    
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := quant50]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_var := var50]
    named_phenometric_summary <- named_phenometric_summary[, upper_var_bound := upper_bound]
    named_phenometric_summary <- named_phenometric_summary[, lower_var_bound := lower_bound]
      
      # Store results in a list
      result_list[[paste0(name, "_", test_bin_year)]] <- named_phenometric_summary
      
    } # end of else statement 
    } # end of i loop
  } # end of w loop

} # end of j loop

# Combine results into a data.table or other appropriate structure
final_result <- do.call(rbind, result_list)

# Write final result to CSV
fwrite(final_result, "/path/to/final_result.csv")

```








```{r}
# PHENESSE for-loop
for(i in 1:length(unique_names_in_bin_v)){
  i <- 16 # Remove later
  named_subset <- test_bin[scientificName %in% unique_names_in_bin_v[i]]
  if(nrow(named_subset) <= 7){
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID), number_of_records = nrow(named_subset), fifieth_percentile = NA)
  }else{
    # Using kde1
    kde.sims <- kde1d(x = named_subset$doy, xmin = 0, xmax = 365)
    quant <- qkde1d(p = 0.50, obj=kde.sims)
    # Using PHENESSE
    named_phenometric_summary <- data.table(scientificName = unique(named_subset$scientificName), GID = unique(named_subset$GID),
                                            number_of_records = nrow(named_subset))
    
    fiftieth_est <- tryCatch(phenesse::weib_percentile(observations = named_subset$doy, percentile = 0.5, iterations = 500),
                                error = function(e) NA)
    fiftieth_ci <- tryCatch(phenesse::weib_percentile_ci(observations = named_subset$doy, iterations = 10, percentile = 0.50, bootstraps = 100),
                                error = function(e) NA)
    
    x <- weib_percentile_ci(observations = named_subset$doy, iterations = 100,
 percentile = 0.50, bootstraps = 100)
    
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_percentile := fiftieth_est]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_nonParametric_bs_percentile := fiftieth_ci[1]]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_nonParametric_bs_lowerCI := fiftieth_ci[2]]
    named_phenometric_summary <- named_phenometric_summary[, fiftieth_nonParametric_bs_upperCI := fiftieth_ci[3]]

      } # End of else statement
  
} # End of for-loop
```








Using kde1d package, pull out the quantiles from the decadal binned data
```{r}

test <- nested_list_holder_b[[1]][[16]]
kde <- density(test$doy)
kde.sims <- kde1d(x = test$doy, xmin = 0, xmax = 365) # Estimate 
plot(kde.sims)
quant50 <- qkde1d(p=0.50, obj=kde.sims)

```



